---
title: 'Rhino immobilisation in the field: Arterial PO2'
author: "Peter Kamerman"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
output:
  html_document:
    highlight: pygments
    theme: yeti
    number_sections: true
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: true
# For generating supplementary material pdfs
#   pdf_document:
#     highlight: pygments
#     toc: yes
#     toc_depth: 4
#     number_sections: true
# geometry: 
#   - a4paper
#   - margin=25mm
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(influence.ME)
library(car)
library(lme4)
library(lmerTest)

# knitr options
knitr::opts_chunk$set(warning = TRUE,
                      message = TRUE,
                      fig.align = 'center',
                      fig.path = './figures/PO2_field/',
                      echo = TRUE)

# Load data
df <- read_rds('./data/PO2_field.rds')
```

# Quick look at the data
```{r quick_look}
# Dataframe dimensions
dim(df)
# Column names
names(df)
# Top six rows of data
head(df)
# Structure of the dataframe
glimpse(df)
```

# Plot
```{r plot}
# Generate a summary df
df %>%
    # Calculate mean resp rate
    group_by(time_min, intervention) %>%
    summarise(mean = round(mean(PO2), 1)) %>%
    # Plot 
    ggplot(data = .) +
    aes(y = mean,
        x = time_min,
        colour = intervention) +
    geom_point(size = 3) +
    geom_line() +
    scale_colour_brewer(name = 'Intervention: ',
                        type = 'qual', 
                        palette = 'Dark2') +
    labs(title = 'Mean arterial PO2 in rhinos during immobilisation',
         subtitle = '(First measurement made five minutes after recumbancy)',
         x = 'Minutes since recumbancy',
         y = 'Arterial PO2 (mm Hg)') +
    theme_bw(base_size = 14) +
    theme(legend.position = 'top')
```

\newpage

# Analyses

****

**Note**: Analysis compares _Butorphanol_ to _Butorphanol + M5050_, and only includes main effects. Also, unlike the boma trials, all interventions included supplemental oxygen.

****

## Linear mixed models
### With supplemental oxygen
#### Model 
```{r model}
# Build model 
## Used a square-root transform on 'PO2' to normalise the residuals
mod_1 <- lmer(sqrt(PO2) ~ time_min + intervention + (1 | id), 
              data = df)

# Perform type II Anova checking main effects only
anova(mod_1, type = 2)

# Model summary
summary(mod_1)
```

#### Model diagnostics
```{r diagnostics}
# Diagnostics on mod_1
## qq plot
qqPlot(x = resid(mod_1), 
       distribution = 'norm',
       main = 'Check for normality of residuals')
## residual plots
plot(mod_1,
     main = 'Check for homoskedasticity')
plot(mod_1, resid(., type = 'pearson', scaled = TRUE) ~ fitted(.),
     main = 'Check for homoskedasticity\n(standardized residuals)')
## Outliers (cut-off dfbetas > 1)
### Generate influence object
mod_influence <- influence(model = mod_1,
                           group = 'id')
### Plot dfbetas
plot(x = mod_influence, 
     which = 'dfbetas',
     abs = TRUE,
     cutoff = 1)
## Linearity
### Time
plot(x = df$time_min, 
     y = residuals(mod_1, type = 'pearson'),
     main = 'Check for linearity: Time') 
abline(h = 0)
### Intervention
plot(x = df$intervention, 
     y = residuals(mod_1, type = 'pearson'),
     main = 'Check for linearity: Intervention') 
abline(h = 0)
```

The QQ-plot is acceptable, but the residuals vs fitted plot includes some large residuals at the upper end of the fitted values, and therefore we have repeated the analysis using robust regression, as implemenented by the `robustlmm::rlmer` function. 

**Note:** that the `robustlmm::rlmer` function does not produce _p-values_. As such, statistical significance has been assessed by calculating simple percentile bootstrap 95% confidence limits for the beta coefficients, and confidence intervals that did not include 0 were deemed significant.

## Robust regression
```{r setup_2}
# Load package
library(robustlmm)

# Generate a function to generate a bootstrap sample of regression coefficients
boot_coef <- function(d, coeff = 1){
    # Nest dataframe by id 
    # (to ensure that each animal's measurements over time are kept together)
    foo <- tidyr::nest(dplyr::group_by(d, id))
    # Sample rows with replacement (8 rhinos used)
    bar <- sample_n(foo, size = 22, replace = TRUE)
    # Unnest sampled rows
    baz <- tidyr::unnest(bar) 
    # Run model
    mod <- robustlmm::rlmer(PO2 ~ time_min + intervention + (1 | id), 
                            data = baz)
    # Extract desired coefficient
    summary(mod)$coefficients[[coeff]]
}
```

### With supplemental oxygen
#### Model 
```{r robust}
# Build model 
mod_1r <- rlmer(PO2 ~ time_min + intervention + (1 | id), 
                data = df)

# Model summary
summary(mod_1r)
```

#### Calculate confidence intervals
```{r robust_ci_O2}
# Set random seed (for reproducibility)
set.seed(1234)

# Extract point estimate for each coefficient
mod_1_obs_intercept <- summary(mod_1r)$coefficients[[1]]
mod_1_obs_time_min<- summary(mod_1r)$coefficients[[2]]
mod_1_obs_m5050 <- summary(mod_1r)$coefficients[[3]]

# Create 1000 repliactes of each function call for all four coefficients
# (Takes forever to run, so set random seed and ran code outside of knitr, 
# saving the output to .rds files in `robust_regression_cache`, and reading 
# the .rds files into memory when the code chunk is run)
## Intercept
# mod_1_intercept <- replicate(n = 1000,
 #                             expr = boot_coef(d = df, coeff = 1))
  ## Write to file (to save time later)
    # write_rds(mod_1_intercept,
     #          './robust_regression_cache/PO2_field_mod_1_intercept.rds')
     ## Read in saved bootstrap data
     mod_1_intercept <-
         read_rds('./robust_regression_cache/PO2_field_mod_1_intercept.rds')
    
## Time_min
# mod_1_time_min <- replicate(n = 1000,
 #                         expr = boot_coef(d = df, coeff = 2))
  ## Write to file (to save time later)
    # write_rds(mod_1_time_min,
     #          './robust_regression_cache/PO2_field_mod_1_time_min.rds')
     ## Read in saved bootstrap data
     mod_1_time_min <- 
         read_rds('./robust_regression_cache/PO2_field_mod_1_time_min.rds')

## Butorphanol vs Butorphanol + M5050
# mod_1_m5050 <- replicate(n = 1000,
 #                         expr = boot_coef(d = df, coeff = 3))
  ## Write to file (to save time later)
    # write_rds(mod_1_m5050, 
     #          './robust_regression_cache/PO2_field_mod_1_m5050.rds')
     ## Read in saved bootstrap data
     mod_1_m5050 <- 
         read_rds('./robust_regression_cache/PO2_field_mod_1_m5050.rds')

# Calculate simple percentile bootstrap 
## Intercept
quantile(mod_1_intercept,
         p = c(0.975, 0.025))
## Time_min
quantile(mod_1_time_min,
        p = c(0.975, 0.025))
## M5050
quantile(mod_1_m5050,
        p = c(0.975, 0.025))
```

#### Model diagnostics
```{r robust_diagnostics_O2}
# Plots
## Darker color indicates observations with a low robustness weight.
plot(mod_1r)
```

# Session information
```{r session_info}
sessionInfo()
```